
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true

## Loading and preprocessing the data

```{r}
unzip("activity.zip") 
act <- read.csv("activity.csv") 
str(act)
summary(act)
```


## What is mean total number of steps taken per day?

```{r}
# calculate total, mean and median steps by date
library(dplyr)
steps_per_day <- act %>%
        group_by(date) %>%
        summarize(total_steps = sum(steps), 
                  mean_steps = mean(steps, na.rm = TRUE),
                  med_steps = median(steps, na.rm = TRUE)
        ) 

steps_per_day$mean_steps 
```

Make a histogram of the total number of steps taken each day

```{r}
hist(steps_per_day$total_steps, breaks = 20,
     xlab = "Total steps per day", main = "Histogram of total steps per day")
```

Report the mean and median of the total number of steps taken per day

```{r}
steps_per_day$mean_steps
steps_per_day$med_steps
```


## What is the average daily activity pattern?

```{r}
by_interval <- act %>%
        group_by(interval) %>%
        summarise(mn_stp = mean(steps, na.rm = TRUE)) # calculate mean steps for each interval

with(by_interval, plot(interval, mn_stp, type = "l", 
                       ylab = "Average number of steps",
                       xlab = "Interval of day", 
                       main = "Average no. of steps across all days by 5-minute interval"))
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
which.max(by_interval$mn_stp)
```

## Imputing missing values

```{r}
sum(is.na(act))
```

Fill in all of the missing values in the dataset with the mean for that 5-minute interval.

```{r}
mn_int <- tapply(act$steps, act$interval, mean, na.rm = TRUE) # calculation of mean steps per interval
act$step.narm <- replace(act$steps, is.na(act$steps), mn_int) # new "steps" variable in data set with NA values replaced by the mean at each interval
```

Make a histogram of the total number of steps taken each day with the imputed values

```{r}
steps_per_day_narm <- act %>%
  group_by(date) %>%
  summarize(total_steps = sum(step.narm),
            mean_steps = mean(step.narm),
            med_steps = median(step.narm)
  ) ## calculate total, mean and median steps per day with imputed NA values

par(mfrow = c(2, 1)) ## make plots side by side for comparison
hist(steps_per_day_narm$total_steps, breaks = 20, 
     xlab = "Steps per day (removed NAs)",
     main = "Histogram with NAs imputed by mean per interval")
hist(steps_per_day$total_steps, breaks = 20,
     xlab = "Steps per day (including NAs)",
     main = "Historgram includes NAs")
```

Recalculate and report the mean and median total number of steps taken per day.

```{r}
steps_per_day_narm$mean_steps
steps_per_day_narm$med_steps
```

Do these values differ from the estimates from the first part of the assignment? 

```{r}
summary(steps_per_day$mean_steps)
summary(steps_per_day_narm$mean_steps)
summary(steps_per_day$med_steps)
summary(steps_per_day_narm$med_steps)
## make figures for comparison
library(ggplot2)
library(cowplot)
p1 <- ggplot(steps_per_day, aes(as.numeric(date), mean_steps)) +
    geom_line() +
    labs(x = "Day", y = "Mean no. of steps", title = "Avg steps per day (include NAs)")

p2 <- ggplot(steps_per_day_narm, aes(as.numeric(date), mean_steps)) +
    geom_line() + 
    labs(x = "Day", y = "Mean no. of steps", title = "Avg steps per day (NAs imputed)")

p3 <- ggplot(steps_per_day, aes(as.numeric(date), med_steps)) +
    geom_line() +
    labs(x = "Day", y = "Median no. of steps", title = "Median steps per day (include NAs)")

p4 <- ggplot(steps_per_day_narm, aes(as.numeric(date), med_steps)) +
    geom_line() +
    labs(x = "Day", y = "Median no. of steps", title = "Median steps per day (NAs imputed)")

plot_grid(p1, p2, p3, p4, ncol = 2)
```

## Are there differences in activity patterns between weekdays and weekends?

```{r}
library(timeDate)
act$date <- as.Date(act$date)  # convert date variable to data format
act$day <- weekdays(act$date) # create day variable

#subset data and add variable for weekend/weekday
wkends <- act %>%
  filter(day == "Sunday"| day == "Saturday") %>%
  mutate(wkday = "weekend")

wkdys <- act %>%
  filter(day == "Monday"| day == "Tuesday" | day == "Wednesday" | 
           day == "Thursday" | day == "Friday") %>%
  mutate(wkday = "weekday")

all_dat <- bind_rows(wkends, wkdys) %>%
        arrange(date) ## new data set with weekend and weekday
```

Panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.

```{r}
# calculate mean steps for weekday
by_int_wkdy <- wkdys %>%
  group_by(interval) %>%
  summarise(mn_stp = mean(step.narm)) 

# calculate mean steps for weekend
by_int_wkend <- wkends %>%
  group_by(interval) %>%
  summarise(mn_stp = mean(step.narm)) 

par(mfrow = c(2, 1))

with(by_int_wkdy, plot(interval, mn_stp, type = "l", 
                       ylab = "Average number of steps",
                       xlab = "Interval of day", 
                       main = "Average no. of steps across weekdays by 5-minute interval"))
with(by_int_wkend, plot(interval, mn_stp, type = "l", 
                       ylab = "Average number of steps",
                       xlab = "Interval of day", 
                       main = "Average no. of steps across weekend by 5-minute interval"))
```

